Parsing Postfix log files

### 2007-04-08 doc/logparser.tex
Lots of corrections from proof-reading. Added a motivation section, needs to be expanded. Expanded the abstract and other sections with uses for the gathered data; this should make it blindingly obvious that the data is useful and the whole thing is more than an academic exercise. Added regex expansions. Added acceptable variable names,and enforced them in the parser. Added a sample for log line for the example rule. Ordered actions so they correspond with the order in which they're usually called. Some more limitations and improvements. Expanded on lots of things, clarified too.

% Better typesetting of URLs.
\usepackage{url}
% Include images
\usepackage[final]{graphicx}

@@

    Parsing Postfix logs is much more difficult than it first appears but
    it is possible.  This paper describes the process required, documenting
    the parsing algorithm and rules, explaining the difficulties
    encountered, with reference to an implementation which stores data
    gathered from the logs in an SQL database for later processing.

@@

@@

\section{Introduction}

Most mail server administrators will have performed some basic processing
of Postfix logs at one time or another, whether it was to debug a problem,
explain to a user why their mail is being rejected, or to check whether
their new anti-spam measures are working.  The more adventurous will have
generated some statistics to show how many hits each of their anti-spam
measures has gotten in the last week, and possibly even generated some
graphs to clearly illustrate the point to management or complaining
users.\footnote{This was the author's first real foray into processing
Postfix logs.}  Very few will have performed in-depth parsing and analysis
of their logs, where the parsing must correlate the log lines
per-connection or per-queueid, rather than processing lines independently.
One of the barriers to this type of processing is the unstructured nature
of Postfix's logs, where each logging line was added on an ad hoc basis as
a requirement was discovered or new functionality was added.  Further
complication arises as a result of the facility to have rejection messages
defined by the administrator, and every Real-time Black List returns a
different explanatory message.  This paper documents the difficult process
of parsing Postfix logs, and presents a program which parses logs and
places the resulting data into a database for further processing by other
applications.

@@

@@

Postfix is a highly configurable, high performance, secure and scalable
Mail Transport Agent.  It features extensive anti-spam restrictions,
allowing a mail administrator to deploy those restrictions which they judge
suitable for their needs, rather than a fixed set chosen by Postfix's
author.  These restrictions can be selectively applied, combined and
bypassed on a per-client, per-recipient or per-sender basis, allowing
varying levels of defense and or permissiveness.  Postfix leverages simple
lookup tables to support arbitrarily complicated user-defined sequences of
restrictions and exceptions, with the ultimate in flexibility being the
facility to consult an external process which can implement whatever logic
is required: some users can be restricted to sending mail on the third
Tuesday after pay day only, if that is of any use.\footnote{This example
may actually be useful in a payroll system.  More commonly encountered
scenarios are restricting posting to large mailing lists (e.g.\ sending
weekly offers to customers) to out of office hours, rate limiting clients,
or checking SPF records (see http://www.openspf.org/ or
http://en.wikipedia.org/wiki/Sender\_Policy\_Framework for details)}
Administrators can also supply their own rejection messages, to make it
clear to senders why exactly their mail was rejected.  Unfortunately this
flexibility has a cost: complexity in the logs generated.  While it is easy
to use \texttt{grep(1)} to determine the fate of an individual email,
following the journey an email takes through Postfix can be quite
difficult.  The logs tend to follow a 90\%-10\% pattern: 90\% of the time
the journey is simple, but the other 10\% of the time requires 90\% of the
code.\footnote{These numbers don't have a solid scientific basis, they're
based on gut feeling from writing and debugging the software.}

\subsection{Paper background}

This paper and the program it describes are part of a larger project to
optimise a server's Postfix restrictions, generate statistics and graphs,
and provide a platform on which new restrictions can be trialled and
evaluated to see if they are worth using in the fight against spam.  The
program parses Postfix logs and populates a database with the data gleaned
from those logs, providing a consistent and simple view of the logs which
future tools can utilise.  The program should be reasonably easily
extensible, so if logging to a database doesn't meet requirements,
appropriate functionality can be implemented.

@@

Before getting into detail about the parser a brief overview is in order.
The parser is split into two parts: the parsing algorithm and the rules
which are applied to the lines.  Rules can be thought of as the Lex part of
a Lex and YACC style parser; rules identify the line and return some data,
which the algorithm (the YACC part) receives, performing the requested
action.  Rules are solely concerned with identifying a line and extracting
data from it, whereas the algorithm's task is to follow the journey each
mail takes through Postfix, piecing the data together into a coherent
whole, and performing housekeeping duties.

@@

    \item The logs are whole and complete; nothing has been removed, either
        deliberately or accidentally (e.g.\ log rotation gone awry, file
        system filling up, logging unable to cope with the volume of logs).

@@

In some ways this task is similar to reverse engineering a program, or
replicating a black box program based solely on its inputs and outputs.
Although the source code is available, reading and understanding it would
require a significant investment of time:

\begin{tabular}[]{lll}

    Postfix 2.3.8   & Postfix 2.4.0 &                   \\
    82224           & 83965         & lines of code     \\
    67146           & 68675         & lines of comments \\
    17647           & 18069         & lines are blank   \\
    167017          & 170709        & lines in total    \\

@@

@@

The complexity and variation in Postfix's logs requires similar flexibility
in the parser; decoupling the parsing rules from the associated actions
allows new rules to be written and tested without requiring modifications
to the algorithm source code (significantly lowering the barrier to entry
for new or casual users), and greatly simplifies both algorithm and rules.
It also allows for a clear separation of functionality: rules handle low
level details of identifying and extracting data from a line, whereas the
algorithm handles the higher level details of following the path a mail
takes through postfix, assembling the required data, etc.

@@

    \item The first matching rule wins: no further rules are tried against
        that line,\footnote{With one exception, details are provided later
        in Section~\ref{actions-in-detail}.} but there is a facility for
        the specifying the order of the rules, so that more specific rules
        can be tried first.

    \item Rules are completely self-contained: there are no sub-rules, so
        rules have linear complexity, and can be understood in isolation,
        without reference to any other rules.

    \item XXX EXPAND ON THIS SOME MORE XXX

@@

    \item [regex] The regex to match the log line against.  The regex will
        first have several keywords expanded: this simplifies reading and
        writing rules, avoids needless repetition of complex regex
        components, and allows the components to be corrected and/or
        improved in one location.  For efficiency the keywords are expanded
        and every rule's regex is compiled before attempting to parse the
        log file --- otherwise each regex would be recompiled each time it
        was used, resulting in an 468.95\% slowdown.

@@

            \item Compiled run time was 55.47 seconds, standard deviation
                was 0.247 seconds.

            \item Non-compiled run time was 260.14 seconds (4 minutes,
                20.14 seconds), standard deviation was 0.96 seconds.

@@

    \item [result\_cols, connection\_cols] Specifies how the fields in the
        log line will be extracted.  The format is: \newline
        client\_hostname = 1; recipient = 2; sender = 4; \newline
        i.e. semi-colon separated assignment statements, with the variable
        name on the left and the matching field from the regex on the right
        hand side.

    \item [result\_data, connection\_data] Sometimes rules need to fake a
        piece of data which isn't present in the log line: e.g.\ setting
        smtp\_code when mail is accepted.  The format is the same as
        result\_cols and connection\_cols, except that arbitrary data can
        be specified rather than just a number.

@@

        Uninteresting lines are parsed rather than ignored completely so
        that any lines the parser isn't capable of handling become
        immediately obvious.

@@

            \item Ascending order run time was 56.684 seconds, standard
                deviation was 0.29 seconds.

            \item Descending order (reversed) run time was 64.037 seconds,
                standard deviation was 0.16 seconds.

            \item Shuffled order run time was 57.19 seconds, standard
                deviation was 0.95 seconds.

@@

@@

XXX CARL SUGGESTS EXPANDING ON THIS SOME MORE, BUT I DON'T KNOW HOW\@.
PERHAPS WHEN THE ALGORITHM IS EXPLAINED\@? XXX

@@

While the rules are more than double (treble with 2.3.x rules?) the size of
the algorithm, the rules are quite simple and each rule is completely
independent of its fellows.  The parser is significantly more complicated,
and highly internally interdependent.

\subsection{Naive approach}

@@

    \item Follow the progress of the accepted mail until it's either delivered or
        bounced, log all data, and finish.

@@

@@

            \item Two data structures to hold the data structures for each
                connection.

@@

    \item Multiple independent mails may be delivered during one
        connection: this requires cloning the current data, so that
        subsequent mails won't trample over each other.  This must be done
        every time a mail is accepted, as it's impossible to tell in
        advance which connections will accept multiple mails.  It is quite
        easy to overlook this complication because only a small minority of
        connections accept more than one mail. Happily once the mail has
        been accepted log entries won't be correlated by pid for that mail
        any more (queueid will be used instead), so there isn't any
        ambiguity about which mail a given log line belongs to.  The
        original connection will be discarded unsaved when the client
        disconnects.  One unsolved difficulty is distinguishing between
        different groups of rejections, e.g.

@@

    \item The most difficult complication is that mails are not always
        delivered directly to a mailbox (or program; there is very little
        difference between the two): sometimes they are aliased and
        need to be delivered to an address (or addresses) on another
        server.  When this occurs a child mail will be injected into the
        postfix queue, but without the explicit logging smtpd or sendmail
        injected mails have, so the source is not immediately discernible;
        from a strictly linear reading of the logs it \textit{usually\/}
        appears as if the child mail has appeared from thin air.
        Subsequently the parent mail will log the creation of the child
        mail:

@@

        This has the effect that the parser cannot complain when it
        encounters a log line from qmgr for a previously unseen mail; it
        must mark the mail as potentially fake and subsequently clear the
        faked flag if and when the origin of the mail becomes clear.
        Obviously the parser could omit checking of where mails originate
        from, but I feel that it is better to require an explicit source,
        as bugs are more likely to be exposed.

        Process scheduling can have a still more confusing effect: quite
        often the child mail will be created, delivered and removed
        \textbf{before} the parent logs the creation line!  Thus mails
        marked as faked cannot be entered into the database; instead they
        must be marked as database ready and subsequently entered by the
        parent mail.  Quite apart from identifying mail injected in an
        unknown fashion, or bugs in the parser, faked mails need to be
        marked as such because they lack some data present in the parent
        mail, and must copy that data from their parent.  XXX EITHER REMOVE
        THE LAST LINE OR EXPAND ON IT SOME MORE, AND FIGURE OUT WHY MORE
        DATA ISN'T REQUIRED XXX

@@

\subsection{Flow chart}

@@

The intermingling of log entries from different mails immediately rules out
the possibility of handling each mail in isolation; the parser must be
capable of handling multiple mails in parallel, each potentially at a
different stage in its journey, without any interference between mails ---
except in the minority of cases where intra-mail interference is required.
I felt that the best way to implement this was to maintain state
information for every unfinished mail and manipulate it appropriately for
each log line encountered.  The parser thus requires both a method of
mapping log lines to the correct mail and a method of specifying the action
the log line represents.  The former is achieved by using the pid of the
smtpd during the initial phase, then switching to the queueid once the mail
has been accepted.  The latter uses the action field of the rule which
matched the log line, executing the code in the function named by the
action.  From a high level viewpoint this design shares significant
similarity with how an editor maintains state for multiple files and
performs the actions the user initiates on the correct file.

@@

Section~\ref{actions-in-detail} explains the actions in substantive detail;
this section will omit such detail because it would clutter and confuse the
algorithm description.  The flow chart in section~\ref{flow-chart} should
be consulted also.

@@

Everything starts off with a mail entering the system, by local submission
via postdrop/sendmail, by SMTP, or by re-injection due to forwarding.
Local submission is the simplest of the three: a queueid is assigned
immediately and the sender address is logged (action: pickup;
flowchart:~2).  SMTP is more complicated:

@@

    \item This is followed either by rejections (action: save\_by\_pid;
        flowchart:~4) or acceptance of one or more mails (action: clone,
        then save\_by\_pid; flowchart:~5,~4), or a combination of both.

    \item The client disconnects (action: disconnect; flowchart:~6).  As
        soon as a mail is accepted it is assigned a queueid; if there is no
        queueid Postfix didn't accept any mail over that connection, it
        rejected all attempts due to the configured restrictions.  Absence
        of a queueid is checked for during disconnection handling (action:
        still disconnect; flowchart:~7) and if none is found the data is
        cleaned up and entered in the database (action: still disconnect;
        flowchart:~8; subroutines: fixup\_connection and
        commit\_connection), then removed from memory (action: still
        disconnect; flowchart:~9).

    \item If one or more mails were accepted there will be more log entries
        later.

@@

Re-injection due to forwarding sadly lacks explicit log lines of its own;
there are two implicit indications:

@@

Re-injection is somewhat awkward to explain because it overlaps both the
mail entering and mail delivery sections.  See
section~\ref{tracking-re-injected-mail} for a full discussion.

The obvious counterpart to mail entering the system is mail leaving the
system, whether it is by deletion, bouncing or delivery.  All three are
handled in exactly the same way:

    \item Eventually the mail will be delivered (even if re-injected),
        bounced, or deleted by the administrator (action: commit;
        flowchart:~12).  This is the last line in the logs for this
        particular mail (it may be indirectly referred to if it was
        re-injected).  If it is neither parent nor child of re-injection
        the data is cleaned up and entered in the database (action: still
        commit; flowchart:~14), then deleted from memory (action: still
        commit; flowchart:~15).  For re-injected mails see 
        section~\ref{tracking-re-injected-mail}.

It's probably apparent by now that tracking re-injected mails is the single
most complex part of the parser.

@@

@@

    \item [ignore] Ignore this line, so that all log lines we're interested
        in are parsed, but this line doesn't have any effect.

@@

Every piece of software suffers from some limitations, and there is almost
always room for improvement.

@@

    \item Each new Postfix release requires new rules to be written, to
        cope with the new messages.

    \item The program should save state at the end of each run, and reload
        it at the start of the next run, to properly cope with mails which
        span logs.  This might also require purging old mails once they've
        been in the state tables for too long, though the presence of very
        old mails indicates a bug in either the implementation or the
        specification.

    \item It appears that the hostname used in the HELO command is not
        logged if the mail is accepted.\footnote{Tested with Postfix
        2.2.10; this may possibly have changed in Postfix 2.3, or the
        upcoming 2.4.}  It should be reasonably simple to write a policy
        server which causes Postfix to log a warning containing the HELO
        hostname when the DATA command is accepted.

@@

@@

    \item Write the policy server referred to in the third limitation
        above.

@@

>>>>

% Better typesetting of URLs.
\usepackage{url}
% Include images
\usepackage[final]{graphicx}
\renewcommand{\labelenumii}{\roman{enumii}:}

@@

    Parsing Postfix logs is much more difficult than it first appears but
    it is possible.  This paper describes the process required, documenting
    the parsing algorithm and rules, explaining the difficulties
    encountered, with reference to an implementation which stores data
    gathered from the logs in an SQL database.  The gathered data can then
    be used to optimise current anti-spam measures, provide a baseline to
    test new anti-spam measures against, or to produce statistics showing
    how effective those measures are.

@@

XXX FIGURE OUT WHEN TO USE MAIL AND WHEN TO USE CONNECTION; POSSIBLY STATE
THAT THEY'RE MORE OR LESS EQUIVALENT  XXX

@@

\newpage
\section{Introduction}

Most mail server administrators will have performed some basic processing
of Postfix logs at one time or another, whether it was to debug a problem,
explain to a user why their mail is being rejected, or to check whether
their new anti-spam measures are working.  The more adventurous will have
generated statistics to show how many hits each of their anti-spam measures
has gotten in the last week, and possibly even generated some graphs to
clearly illustrate the point to management or complaining
users.\footnote{This was the author's first real foray into processing
Postfix logs.}  Very few will have performed in-depth parsing and analysis
of their logs, where the parsing must correlate the log lines
per-connection or per-queueid rather than processing lines independently.
One of the barriers to this type of processing is the unstructured nature
of Postfix logs, where each logging line was added on an ad hoc basis as a
requirement was discovered or new functionality was added.  Further
complication arises as a result of the facility to have rejection messages
defined by the administrator, every Real-time Black List returns a
different explanatory message, policy servers may have different messages
depending on the characteristics of the connection, \ldots.  This paper
documents the difficult process of parsing Postfix logs, and presents a
program which parses logs and places the resulting data into a database.
The gathered data can then be used to optimise current anti-spam measures,
provide a baseline to test new anti-spam measures against, or to produce
statistics showing how effective those measures are.  There are numerous
other uses for such data: improving server performance by identifying
troublesome destinations and reconfiguring appropriately or dedicating
transports to those destinations; identifying high volume uses (e.g.
customer newsletters) and restricting those uses to off-peak times; as a
base for billing customers on a shared server; etc.

XXX MAYBE EXPLAIN WHAT WILL BE IN THE FOLLOWING SECTIONS??? XXX

@@

\subsection{Motivation}

This paper and the program it describes are part of a larger project to
optimise a server's Postfix restrictions, generate statistics and graphs,
and provide a platform on which new restrictions can be trialled and
evaluated to see if they are worth using in the fight against spam.  The
program parses Postfix logs and populates a database with the data gleaned
from those logs, providing a consistent and simple view of the logs which
future tools can utilise.

XXX EXPLAIN WHY THIS IS USEFUL AND IMPORTANT XXX

XXX ADD A SECTION EXPLAINING THE IMMEDIATE BENEFITS \\
\texttt{SELECT name, description, restriction\_name, rule\_order \\
    FROM rules \\
    ORDER BY rule\_order;}\\
Tells you which of your restrictions are most effective.

XXX COMPARE TO PREVIOUS WORK XXX

@@

Postfix is a highly configurable, high performance, secure and scalable
Mail Transport Agent.  It features extensive anti-spam restrictions,
allowing a mail administrator to deploy those restrictions which they judge
suitable for their needs, rather than a fixed set chosen by Postfix's
author.  These restrictions can be selectively applied, combined and
bypassed on a per-client, per-recipient or per-sender basis, allowing
varying levels of defense and or permissiveness.  Postfix leverages simple
lookup tables to support arbitrarily complicated user-defined sequences of
restrictions and exceptions, with the ultimate in flexibility being the
facility to consult an external process which can implement whatever logic
is required: some users can be restricted to sending mail on the third
Tuesday after pay day only, if that is of any use.\footnote{This example
may actually be useful in a payroll system.  More commonly encountered
scenarios are restricting posting to large mailing lists (e.g.\ sending
weekly offers to customers) to out of office hours, rate limiting clients,
or checking SPF records (see http://www.openspf.org/ or
http://en.wikipedia.org/wiki/Sender\_Policy\_Framework for details)}
Administrators can also supply their own rejection messages to make it
clear to senders why exactly their mail was rejected.  Unfortunately this
flexibility has a cost: complexity in the logs generated.  While it is easy
to use \texttt{grep(1)} to determine the fate of an individual email,
following the journey an email takes through Postfix can be quite
difficult.  The logs tend to follow a 90\%-10\% pattern: 90\% of the time
the journey is simple, but the other 10\% of the time requires 90\% of the
code.\footnote{These numbers don't have a solid scientific basis, they're
based on gut feeling from writing and debugging the software.}

@@

Before getting into detail about the parser a brief overview is in order.
The parser is split into two parts: the parsing algorithm and the rules
which are applied to the lines.  Rules can be thought of as the Lex part of
a Lex and YACC style parser; rules identify the line and return some data,
which the algorithm (the YACC part) receives and performs the requested
action.  Rules are solely concerned with identifying a line and extracting
data from it, whereas the algorithm's task is to follow the journey each
mail takes through Postfix, piecing the data together into a coherent
whole, saving it in a useful and consistent form, and performing
housekeeping duties.

@@

    \item The logs are whole and complete; nothing has been removed, either
        deliberately or accidentally (e.g.\ log rotation gone awry, file
        system filling up, logging system unable to cope with the volume of
        logs).

@@

In some ways this task is similar to reverse engineering or replicating a
black box program based solely on its inputs and outputs.  Although the
source code is available, reading and understanding it would require a
significant investment of time:

\begin{tabular}[]{llll}

    Postfix 2.2.11  & Postfix 2.3.8   & Postfix 2.4.0 &                   \\
    71548           & 82224           & 83965         & lines of code     \\
    60962           & 67146           & 68675         & lines of comments \\
    16117           & 17647           & 18069         & lines are blank   \\
    148627          & 167017          & 170709        & lines in total    \\

@@

\subsection{Other parsers}

XXX ADD INFORMATION ABOUT OTHER POSTFIX LOG PARSERS HERE XXX

@@

The complexity and variation in Postfix's logs requires similar flexibility
in the parser; decoupling the parsing rules from the associated actions
allows new rules to be written and tested without requiring modifications
to the algorithm source code (significantly lowering the barrier to entry
for new or casual users), and greatly simplifies both algorithm and rules.
It also creates a clear separation of functionality: rules handle low level
details of identifying and extracting data from a line, whereas the
algorithm handles the higher level details of following the path a mail
takes through Postfix, assembling the required data, etc.

@@

    \item The first matching rule wins: no further rules are tried against
        that line,\footnote{With one exception, details are provided later
        in Section~\ref{actions-in-detail}.} but there is a facility for
        the specifying the order of the rules so that more specific rules
        can be tried first.

    \item Rules are completely self-contained and can be understood in
        isolation, without reference to any other rules.

    \item There are no sub-rules, so rules have linear computational
        complexity.

@@

    \item [regex] The regular expression\footnote{See
        http://en.wikipedia.org/wiki/Regular\_expression for more
        information about regular expressions in general.} to match the log
        line against.  The regex will first have several keywords expanded:
        this simplifies reading and writing rules, avoids needless
        repetition of complex regex components, and allows the components
        to be corrected and/or improved in one location.

        The following keywords are expanded:

        \texttt{\_\_SENDER\_\_, \_\_RECIPIENT\_\_, \_\_HELO\_\_,
        \_\_HOSTNAME\_\_, \_\_IP\_\_, \newline\_\_SMTP\_CODE\_\_,
        \_\_QUEUEID\_\_ and \_\_COMMAND\_\_.}

        \_\_COMMAND\_\_ is an SMTP command, e.g. MAIL FROM; hopefully the
        others are reasonably self explanatory.

        For efficiency the keywords are expanded and every rule's regex is
        compiled before attempting to parse the log file --- otherwise each
        regex would be recompiled each time it was used, resulting in an
        468.95\% slowdown.

@@

            \item [Compiled once and saved:] run time was 55.47 seconds,
                standard deviation was 0.247 seconds.

            \item [Compiled every time and discarded:] run time was 260.14
                seconds (4 minutes, 20.14 seconds) (468.95\% slower),
                standard deviation was 0.96 seconds.

@@

        XXX DO I NEED TO SAY ANYTHING MORE ABOUT THIS??

    \item [result\_cols, connection\_cols] Specifies how the fields in the
        log line will be extracted.  The format is: \newline
        \texttt{client\_hostname = 1; recipient = 2, sender = 4;} \newline
        i.e.\ semi-colon or comma separated assignment statements, with the
        variable name on the left and the matching field from the regex on
        the right hand side.  The list of acceptable variable names is:

        \texttt{client\_hostname, client\_ip, server\_ip, server\_hostname,
        \newline helo, sender, recipient, smtp\_code, data and child.}

    \item [result\_data, connection\_data] Sometimes rules need to supply a
        piece of data which isn't present in the log line: e.g.\ setting
        \texttt{smtp\_code} when mail is accepted.  The format and allowed
        variables are the same as for \texttt{result\_cols} and
        \texttt{connection\_cols}, except that arbitrary
        data\footnote{Commas and semi-colons cannot be escaped and thus
        cannot be used.  This is intended for small amounts of data rather
        than large, so dealing with escape sequences seemed unnecessary.}
        is permitted on the right hand side of the assignment.

@@

        Uninteresting lines are parsed so that any lines the parser isn't
        capable of handling become immediately obvious errors.

@@

            \item [Ascending order (best):] run time was 56.684 seconds,
                standard deviation was 0.29 seconds.

            \item [Descending order (worst):] run time was 64.037
                seconds (12.97\% slower), standard deviation was 0.16
                seconds.

            \item [Shuffled order:] run time was 57.19 seconds
                (0.89\% slower), standard deviation was 0.95 seconds.

        XXX RERUN OVER MORE LOG FILES XXX

@@

        It is hoped that this will prove to have greater effectiveness as
        the number of rules grows; currently the effect is negligible,
        particularly in comparison to compiling each regex once.

@@

This rule would match the following log line:

\begin{verbatim}
NOQUEUE: reject: RCPT from example.com[10.1.1.1]: 
  550 <foo@bar.baz>: Sender address rejected: Domain not found;
  from=<foo@bar.baz> to=<info@example.com> proto=SMTP
  helo=<smtp.bar.baz>
\end{verbatim}

XXX CARL SUGGESTS EXPANDING ON THIS SOME MORE, BUT I DON'T KNOW HOW\@.
PERHAPS WHEN THE ALGORITHM IS PROPERLY EXPLAINED IT'LL BE EASIER\@?  MAKE
REFERENCES TO THE ALGORITHM, EXPLAIN MORE CLEARLY HOW THE FIELDS ARE USED,
HOW THE RULE MATCHES, ETC.  XXX

@@

While the rules are more than double (treble with 2.3.x rules? XXX ADD
2.3.X RULES AND CHECK XXX) the size of
the algorithm, the rules are quite simple and each rule is completely
independent of its fellows.  The algorithm is significantly more
complicated and highly internally interdependent.

\subsection{A naive approach}

@@

    \item Follow the progress of the accepted mail until it's either
        delivered, bounced or deleted, log all data, and finish.

@@

\label{complications}

@@

            \item Two containing data structures to hold the data
                structure for each connection.

@@

    \item Multiple independent mails may be delivered during one
        connection: this requires cloning the current data as soon as a
        mail is accepted, so that subsequent mails won't trample over each
        other's data.  This must be done every time a mail is accepted, as
        it's impossible to tell in advance which connections will accept
        multiple mails.  It is quite easy to overlook this complication
        because only a small minority of connections accept more than one
        mail. Happily once the mail has been accepted log entries won't be
        correlated by pid for that mail any more (its queueid will be used
        instead), so there isn't any ambiguity about which mail a given log
        line belongs to.  The original connection will be discarded unsaved
        when the client disconnects.  One unsolved difficulty is
        distinguishing between different groups of rejections, e.g. when
        dealing with the following sequence:

@@

    \item The most difficult complication is that mails are not always
        delivered directly to a mailbox (or program; there is very little
        difference between the two): sometimes they are accepted for a
        local address but need to be delivered to one or more remote
        addresses due to aliases.  When this occurs a child mail will be
        injected into the postfix queue, but without the explicit logging
        smtpd or sendmail injected mails have, so the source is not
        immediately discernible from the log line in which the mail first
        appears; from a strictly linear reading of the logs it
        \textit{usually\/} appears as if the child mail has appeared from
        thin air.  Subsequently the parent mail will log the creation of
        the child mail:

@@

        Because of this the parser cannot complain when it encounters a log
        line from qmgr for a previously unseen mail; it must mark the mail
        as potentially fake and subsequently clear the faked flag if and
        when the origin of the mail becomes clear.  Obviously the parser
        could omit checking of where mails originate from, but I feel that
        it is better to require an explicit source, as bugs are more likely
        to be exposed.

        Process scheduling can have a still more confusing effect: quite
        often the child mail will be created, delivered and entirely
        finished with \textbf{before} the parent logs the creation line!
        Thus mails marked as faked cannot be entered into the database when
        their final log line is parsed; instead they must be marked as
        database ready and subsequently entered by the parent mail once it
        has been identified.  Quite apart from identifying mail injected in
        an unknown fashion, or bugs in the parser, faked mails need to be
        marked as such because they lack some data present in the parent
        mail, and must copy that data from their parent before being
        entered in the database.  XXX EITHER REMOVE THE LAST LINE OR EXPAND
        ON IT SOME MORE.  FIGURE OUT WHY MORE DATA ISN'T REQUIRED.

        The algorithm requirements are discussed in
        section~\ref{tracking-re-injected-mail}.

@@

\newpage
\subsection{Flow chart}

@@

The intermingling of log entries from different mails immediately rules out
the possibility of handling each mail in isolation; the parser must be
capable of handling multiple mails in parallel, each potentially at a
different stage in its journey, without any interference between mails ---
except in the minority of cases where intra-mail interference is required.
I felt that the best way to implement this was to maintain state
information for every unfinished mail and manipulate the appropriate mail
correctly for each log line encountered.  The parser thus requires both a
method of mapping log lines to the correct mail and a method of specifying
the action the log line represents.  The former is achieved by using the
pid of the smtpd to identify the correct mail during the initial phase,
then switching to the queueid once the mail has been accepted.  The latter
uses the action field of the rule which matched the log line, executing the
code in the function named by the action.  From a high level viewpoint this
design shares significant similarity with how an editor maintains state for
multiple files and performs the actions the user initiates on the correct
file.

@@

Section~\ref{actions-in-detail} explains the actions in substantive detail;
this section will omit such detail because it would clutter and confuse the
algorithm description.  The flow chart in section~\ref{flow-chart} should
be consulted while reading this section also.

@@

Everything starts off with a mail entering the system, whether by local
submission via postdrop/sendmail, by SMTP, or by re-injection due to
forwarding.  Local submission is the simplest of the three: a queueid is
assigned immediately and the sender address is logged (action: pickup;
flowchart:~2).

SMTP is more complicated:

@@

    \item This is followed by rejection of sender, recipients, client
        address, etc. (action: save\_by\_pid; flowchart:~4); acceptance of
        one or more mails (action: clone, then save\_by\_pid;
        flowchart:~5,~4); or some interleaving of both.

    \item The client disconnects (action: disconnect; flowchart:~6).  As
        soon as a mail is accepted it is assigned a queueid; if there is no
        queueid Postfix didn't accept any mail over that connection, it
        rejected all attempted deliveries due to the configured
        restrictions.  Absence of a queueid is checked for during
        disconnection handling (action: still disconnect; flowchart:~7) and
        if none is found the data is cleaned up and entered in the database
        (action: still disconnect; flowchart:~8; subroutines:
        fixup\_connection and commit\_connection), then removed from memory
        (action: still disconnect; flowchart:~9).

    \item If one or more mails were accepted there will be more log entries
        later, see section~\ref{mail-delivery}.

@@

Re-injection due to forwarding sadly lacks explicit log lines of its
own;\footnote{Previously discussed in section~\ref{complications},
complication 3.}  Re-injection is somewhat awkward to explain because it
overlaps both the mail acceptance and mail delivery sections.  See
section~\ref{tracking-re-injected-mail} for a full discussion.

\label{mail-delivery}

The obvious counterpart to mail entering the system is mail leaving the
system, whether it is by deletion, bouncing, local or remote delivery.  All
three are handled in exactly the same way:

    \item Eventually the mail will be delivered (even if re-injected),
        bounced, or deleted by the administrator (action: commit;
        flowchart:~12).  This is the last log line for this particular mail
        (though it may be indirectly referred to if it was re-injected).
        If it is neither parent nor child of re-injection the data is
        cleaned up and entered in the database (action: still commit;
        flowchart:~14; subroutines fixup\_connection and
        commit\_connection), then deleted from memory (action: still
        commit; flowchart:~15).  For re-injected mails see
        section~\ref{tracking-re-injected-mail}.

It's probably apparent by now that tracking re-injected mails is the single
most complex part of the parser.  Section~\ref{complications} complication
3 has most of the previous discussion, should you wish to review it.

there are two implicit indications:

@@

@@

XXX ADD A LIST IN LOAD\_RULES() SPECIFYING THE ALLOWED ACTIONS XXX

This list contains all the actions available in the algorithm, in the order
in which they tend to occur in practice.

@@

    \item [ignore] Do nothing with this line, merely parse it so that
        unparsed lines can be reported.

@@

Every piece of software suffers from some limitations and there is almost
always room for improvement.

@@

    \item Each new Postfix release requires new rules to be written to cope
        with the new messages.  Similarly new RBLs, new policy servers and
        new administrator defined rejection messages require new rules.

    \item The program doesn't cope with mails which span log files:
        currently they are reported as unfinished mails at the end of the
        first run and are complained about when encountered during the next
        run because they don't have a source.

    \item It appears that the hostname used in the HELO command is not
        logged if the mail is accepted.\footnote{Tested with Postfix
        2.2.10; this may possibly have changed in Postfix 2.3, or the
        upcoming 2.4.}  It should be reasonably simple to write a policy
        server which causes Postfix to log a warning containing the HELO
        hostname when the DATA command is accepted.\footnote{See
        http://www.postfix.org/SMTPD\_POLICY\_README.html for full
        documentation about policy servers.}

@@

    \item The program will not detect parsing the same log file twice,
        resulting in the database containing duplicate entries.

@@

    \item Write the policy servers referred to in the limitations above.

    \item Save state at the end of each run; load it at the start of the
        next run.  This should allow the program to cope with mails
        spanning log files.  This might also require purging old mails once
        they've been in the state tables for too long, though the presence
        of very old mails indicates a bug in either the implementation or
        the specification.

@@

\section{References}
